String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-04-28 2023-1-19/sections/introduction.tex
overview

Nevertheless, the aforementioned state-of-the-art string constraint solvers still suffer from the string constraints with regex-counting and string-length, especially when the counting bounds are large. For instance, none of string solvers e.g. CVC4/5, Z3, Z3str3RE, OSTRICH is capable of solving the following constraint within 300 seconds,
%
\begin{equation}\label{eqn-running}
x \in (\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}} \wedge x \in \Sigma^* c \wedge |x| > 120.
\end{equation}
Intuitively, the constraint in (\ref{eqn-running}) specifies that $x$ is a concatenation of three subwords $x_1, x_2, x_3$ where $a$ (resp. $b, c$) does not occur in $x_1$ (resp. $x_2, x_3$), moreover, $x$ ends with $c$, and the length of $x$ is greater than $120$. It is easy to observe that this constraint is unsatisfiable since on the one hand, $|x| > 120$ and the counting upper bound $60$ in  both $(\Sigma \setminus a)^{\{5, 60\}}$ and $(\Sigma \setminus b)^{\{5, 60\}}$ imply that $x$ must end with some character from $\Sigma \setminus c$, that is, a character different from $c$, and on the other hand, $x \in \Sigma^*c$ requires that $x$ has to end with $c$.

>>>>

Nevertheless, the aforementioned state-of-the-art string constraint solvers still suffer from the string constraints with regex-counting and string-length, especially when the counting bounds are large. For instance, none of string solvers e.g. CVC4/5, Z3, Z3str3RE, OSTRICH is capable of solving the following constraint within 300 seconds,
%
\begin{equation}\label{eqn-running}
x \in (\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}} \wedge x \in \Sigma^* c^+ \wedge |x| > 120.
\end{equation}
Intuitively, the constraint in (\ref{eqn-running}) specifies that $x$ is a concatenation of three subwords $x_1, x_2, x_3$ where $a$ (resp. $b, c$) does not occur in $x_1$ (resp. $x_2, x_3$), moreover, $x$ ends with a nonempty sequence of $c$'s, and the length of $x$ is greater than $120$. It is easy to observe that this constraint is unsatisfiable since on the one hand, $|x| > 120$ and the counting upper bound $60$ in  both $(\Sigma \setminus a)^{\{5, 60\}}$ and $(\Sigma \setminus b)^{\{5, 60\}}$ imply that $x$ must end with some character from $\Sigma \setminus c$, that is, a character different from $c$, and on the other hand, $x \in \Sigma^*c^+$ requires that $x$ has to end with $c$.

### 2023-04-28 2023-1-19/sections/overview.tex
overview

@@

However, even when we use CEFA to encode counting in the string constraints, the linear integer arithmetic reduced from the CEFA still needs to be simplified to solve. The reason is that the linear integer arithmetic is solved in exponential time of the number of variables, which is linear to the sum of transitions number and states number in the CEFA. To address this issue, we use symbolic-aware simplification to reduce the number of transitions and states in the CEFA. Simply illustrating our idea, consider a simple NFA rather than a CEFA. We assume the NFA consisted of three states $q_1, q_2, q_3$, where $q_1$ is the initial state, $q_2$ and $q_3$ are the accepting states. Assume we have three transitions from $q_1$ to $q_2$ with different characters, and three transitions from $q_1$ to $q_3$. Actually, we only need to attempt some of the six transitions to get the reachable result. We run any of the transitions, and the reachability is known. So this NFA can be simplified to two states $q_1', q_2'$ where $q_1'$ is the initial state, $q_2'$ is the accepting state, and one transition from $q_1'$ to $q_2'$ with abbreviated character. The size of NFA is sharply reduced. The simplification can be done by treating the character on each transition as the same character, then applying the minimization algorithm to the NFA. CEFA can be simplified similarly, except we need to consider the counting information, i.e., symbolic updates, on each transition. For example, $a:(1)$ and $b:(1)$ are treated as the same label but $a:(1)$ and $b:(0)$ are not. Then we can apply the minimization algorithm as NFA. Recalling the CEFA (Fig. \ref{fig:overview:orgin}) handling the string constraints \ref{eqn-running}, its transitions number between $q_1$ and $q_2$ are decided by the size of alphabet $\Sigma$. We apply symbolic-aware simplification on it, then a much smaller CEFA (Fig. \ref{fig:overview:simplified}) is obtained, whose alphabet is unary so that the number of the transition between $q_1$ and $q_2$ decreases to $1$. The state number does not decrease because these three states are not mutually equivalent. \newline
Sometimes the string constraints are satisfiable, and the strings in the solution have a short length. Such a solution may be quickly explored by derivation rules in DPLL(T)-based string solvers but slowly explored by our approach. To improve efficiency on satisfiable string constraints, we propose a light-way heuristic that tries to find a solution in the under-approximation of the string constraints. The main idea is to explore paths within an exact length and manually compute the registers' values, rather than reduce CEFA to heavy linear integer arithmetic. For example, consider the string constraints $x \in (\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}} \wedge x \in \Sigma^* c$, we can explore paths within length 10 and get a satisfiable solution $x = cccccccccc$. 
\begin{figure}[ht]
  \centering
  \import{figures}{overview_example.tex}
  \caption{The CEFA handling the difficult string constraints}
  \label{fig:overview:orgin}
\end{figure}
\begin{figure}[ht]
  \centering
  \import{figures}{overview_example_simplified.tex}
  \caption{The simplified CEFA handling the difficult string constraints}
  \label{fig:overview:simplified}
\end{figure}

>>>>

In this section, we utilize the string constraint in Equation~(\ref{eqn-running}) to illustrate the approach in this work.

%$x \in (\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}} \wedge x \in \Sigma^* c^+ \wedge |x| > 120$.

\zhilin{epsilon transition should be avoided}

We first construct an automaton for the regular expression $(\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}}$. We introduce three registers, say $r_1, r_2, r_3$, to represent the three counting operators.  Moreover, we introduce another register, say $r_0$, to denote the length of strings. Then we obtain a cost-enriched finite automaton (CEFA) $\aut_1$ that is illustrated in Figure~\ref{fig:overview}(a), where the self-loops around $p_0, p_1, p_2$ represent the first, second, and third counting operator respectively, in which $r_1, r_2, r_3$ are incremented. 
%Moreover, $r_0$ is incremented in all the three self-loops. 
Furthermore, the counting bounds are specified by the accepting condition $5 \le r_1 \le 60 \wedge 5 \le r_2 \le 60 \wedge 0 \le r_3 \le 60$ that is attached to the final state $p_2$. Note here we represent the counting operators symbolically by registers, instead of unfolding them explicitly. Moreover, we also construct a CEFA $\aut_2$ for $\Sigma^* c^+$ (see Figure~\ref{fig:overview}(b)). 
%where the register $r_0$ is used to record the length of strings.

% the register $r_1$ is incremented in the two transitions from $q_0$ to $q_0$ and $q_1$ respectively,  the register $r_2$ is incremented in the two transitions from $q_1$ to $q_1$ and $q_2$ respectively, and the register $r_3$ is incremented in the transition from $q_2$ to $q_2$ itself. Moreover, the register $r_4$ is incremented in each transition.

\begin{figure}[ht]
  \centering
  \includegraphics[width = 0.9\textwidth]{sections/overview-cefa.pdf}
  \caption{CEFA for $(\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}}$ and $\Sigma^* c^+$}
  \label{fig:overview}
\end{figure}

Next, we remove the $\varepsilon$-transitions of $\aut_1$, resulting into $\aut'_1$, then construct $\aut'_1 \times \aut_2$, that is, the product of $\aut'_1$ and $\aut_2$, as illustrated in Figure~\ref{fig:overview:product}. Furthermore, we add one special register, say $r_0$, to record the length of strings. Note that $r_0$ is incremented in all transitions. In $\aut'_1 \times \aut_2$, the accepting condition $5 \le r_1 \le 60 \wedge 5 \le r_2 \le 60 \wedge 0 \le r_3 \le 60 \wedge r_0 > 120$ is attached to both $(p_0,q_1)$ and $(p_1,q_1)$, where $r_0 > 120$ is added as a conjunct to express $|x| > 120$.

\begin{figure}[ht]
  \centering
  \includegraphics[width = 0.5\textwidth]{sections/overview-cefa-product.pdf}
  \caption{Product of $\aut^\prime_1$ and $\aut_2$, where the length register $r_0$ is added}
  \label{fig:overview:product}
\end{figure}

Let $\tau_1, \cdots, \tau_m$ be an enumeration of the transitions of $\aut'_1 \times \aut_2$ and $\myvec{v_1}, \cdots, \myvec{v_m}$ be the vectors of updates on the register values in these transitions. For instance, in the transition from $(p_0, q_1)$ to $(p_1, q_1)$, $r_0$ is incremented, and the values of the other registers is unchanged. Therefore, the vector for this transition is $(1, 0, 0, 0)$. 
Then the satisfiability of the original string constraint is reduced to the nonemptiness of the CEFA $\aut'_1 \times \aut_2$. 
From the result in \cite{}, an LIA formula $\psi(i_1, \cdots, i_m)$ can be computed to express the constraints on the number of occurrences of $\tau_1, \cdots, \tau_m$ in the accepting runs of $\aut'_1 \times \aut_2$, where $i_1, \cdots, i_m$ are the integer variables representing the number of occurrences of $\tau_1, \cdots, \tau_m$ respectively. 
Therefore, the satisfiability of the string constraint in~(\ref{eqn-running}) is reduced to the satisfiability of the following LIA formula
\[
\psi(i_1, \cdots, i_m) \wedge \bigwedge \limits_{0 \le j \le 3} r_j = \sum \limits_{k \in [m]}  i_k v_{k, j+1} \wedge \alpha(r_0, r_1, r_2, r_3),
\]
where for each $k \in [m]$, $v_k = (v_{k,1}, v_{k,2}, v_{k, 3}, v_{k, 4})$ is the vector of updates on the registers $r_0, r_1, r_2, r_3$ in the transition $\tau_k$, and $\alpha(r_0, r_1, r_2, r_3)$ is the accepting condition in $\aut'_1 \times \aut_2$.

\zhilin{stopped here}

\hide{
\begin{figure}[ht]
  \centering
  \import{figures}{overview_example.tex}
  \caption{The CEFA handling the difficult string constraints}
  \label{fig:overview:orgin}
\end{figure}
}

@@

However, even when we use CEFA to encode counting in the string constraints, the linear integer arithmetic reduced from the CEFA still needs to be simplified to solve. The reason is that the linear integer arithmetic is solved in exponential time of the number of variables, which is linear to the sum of transitions number and states number in the CEFA. To address this issue, we use symbolic-aware simplification to reduce the number of transitions and states in the CEFA. Simply illustrating our idea, consider a simple NFA rather than a CEFA. We assume the NFA consisted of three states $q_1, q_2, q_3$, where $q_1$ is the initial state, $q_2$ and $q_3$ are the accepting states. Assume we have three transitions from $q_1$ to $q_2$ with different characters, and three transitions from $q_1$ to $q_3$. Actually, we only need to attempt some of the six transitions to get the reachable result. We run any of the transitions, and the reachability is known. So this NFA can be simplified to two states $q_1', q_2'$ where $q_1'$ is the initial state, $q_2'$ is the accepting state, and one transition from $q_1'$ to $q_2'$ with abbreviated character. The size of NFA is sharply reduced. The simplification can be done by treating the character on each transition as the same character, then applying the minimization algorithm to the NFA. CEFA can be simplified similarly, except we need to consider the counting information, i.e., symbolic updates, on each transition. For example, $a:(1)$ and $b:(1)$ are treated as the same label but $a:(1)$ and $b:(0)$ are not. Then we can apply the minimization algorithm as NFA. Recalling the CEFA (Fig. \ref{fig:overview:orgin}) handling the string constraints \ref{eqn-running}, its transitions number between $q_1$ and $q_2$ are decided by the size of alphabet $\Sigma$. We apply symbolic-aware simplification on it, then a much smaller CEFA (Fig. \ref{fig:overview:simplified}) is obtained, whose alphabet is unary so that the number of the transition between $q_1$ and $q_2$ decreases to $1$. The state number does not decrease because these three states are not mutually equivalent.

Sometimes the string constraints are satisfiable, and the strings in the solution have a short length. Such a solution may be quickly explored by derivation rules in DPLL(T)-based string solvers but slowly explored by our approach. To improve efficiency on satisfiable string constraints, we propose a light-way heuristic that tries to find a solution in the under-approximation of the string constraints. The main idea is to explore paths within an exact length and manually compute the registers' values, rather than reduce CEFA to heavy linear integer arithmetic. For example, consider the string constraints $x \in (\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}} \wedge x \in \Sigma^* c$, we can explore paths within length 10 and get a satisfiable solution $x = cccccccccc$. 
\begin{figure}[ht]
  \centering
  \import{figures}{overview_example.tex}
  \caption{The CEFA handling the difficult string constraints}
  \label{fig:overview:orgin}
\end{figure}
\begin{figure}[ht]
  \centering
  \import{figures}{overview_example_simplified.tex}
  \caption{The simplified CEFA handling the difficult string constraints}
  \label{fig:overview:simplified}
\end{figure}
